A Practical Optional Type System for Clojure

### 2012-12-06 dissertation/intro.tex
Add Supervisor. Revise abstract, motivation and acknowledgements

The initial motivation for implementing an optional type system
for Clojure was an anecdotal account of the struggles
of a Clojure programmer, Jim Duey. In an apparently heroic effort, Duey managed to 
implement a Clojure library for conduits~\footnote{\url{http://www.yesodweb.com/book/conduits}}, 
an advanced form of ``pipes'', using arrows, a generalisation of monads.
Conduits also significantly surpass monads in complexity and are usually reserved
for languages with advanced type systems like Haskell.

@@

Since then, the Clojure community has shown significant interest in this work.
I also developed this project as a Google Summer of Code 2012 project,
after it was selected by the Clojure community as a Clojure Google Summer of Code
project for 2012~\cite{GSOC2012}.
I have also been invited to speak on this project at the Clojure Conj
2012 conference in November, the main international conference related to Clojure.

@@

>>>>

The initial motivation for implementing an optional type system
for Clojure was outlined in a discussion with Jim Duey, a Clojure programmer, at Clojure Conj 2011.
In an apparently heroic effort, Duey managed to
implement a Clojure library for conduits~\footnote{\url{http://www.yesodweb.com/book/conduits}}, 
an advanced form of ``pipes'', using arrows, a generalisation of monads.
Conduits also significantly surpass monads in complexity and are usually reserved
for languages with advanced type systems like Haskell.

@@

The proliferation of other abstract programming styles, like monads, in statically typed languages
also suggested that such a static type system could
help the programmer write and debug such code, while improving confidence that their code is correct.

\subsection{Why does not Clojure feature static type checking?}

Given the advantages of static type checking outlined in the previous section, it is natural to ask:
why does Clojure not already support static type checking?
The simplest answer is that Clojure was not designed to support a static type system.
Instead, features and idioms that are not necessarily compatible with traditional type systems
were favoured, like variable-arity functions, untyped hash-maps used like objects, and extension of program behaviour
at runtime.

Without specifically speaking for the designers of Clojure, at the time Clojure was designed
it was not clear how to reconcile Clojure's feature-set with static type checking in a satisfying way.
Since then, Typed Racket, which add static type checking to the Racket programming language,
have matured and gained acceptance.
Once this occurred, it seemed like only a matter of time until the ideas developed for the Typed Racket project 
were tested for their applicability to Clojure. 
As far as we know, this project is the first serious attempt to build an optional static type system for Clojure.

@@

\subsection{What is the community reaction to Typed Clojure?}

Since I began developing Typed Clojure, the Clojure community has shown significant interest in this work:

\begin{itemize}
  \item I developed this project as a Google Summer of Code 2012 project,
        after it was selected by the Clojure community as a Clojure Google Summer of Code
        project for 2012~\cite{GSOC2012}.
  \item I gave a talk on this project at the Clojure Conj
        2012 conference, the main international programming conference related to Clojure, in November.
\end{itemize}

### 2012-12-06 dissertation/paper.tex
Add Supervisor. Revise abstract, motivation and acknowledgements

\keywords{Type systems, Clojure}
% TODO What are categories? cshonours requires them - but commented out now
%\categories{}

@@

Traditionally there has been a clear distinction between 
programming languages that perform type checking at compile-time, and those that do not.
In particular, the early wide-spread programming languages ALGOL and Lisp
could be divided this way: ALGOL performs compile-time (or static) type checking,
and Lisp does not, and most recent languages can be similarly classified.
Static type checking eliminates many common user errors that are otherwise unnoticed or are caught 
when the program is run.
On the other hand, a language without static type checking can often be designed 
to be more flexible and forgiving, making it easy, for example, to have a \emph{map} function that can map functions
with any number of arguments (as in most Lisp dialects).

A recent trend is to aim to combine the advantages of both kinds of languages by adding optional static 
type systems to languages without static type checking. 
A major challenge in these combinations is continuing to support the idiomatic style of the original 
language while performing accurate enough checking to catch similar errors to standard statically
typed languages. Some recent attempts appear to have finally produced type systems that have overcome this obstacle.
Notably Typed Racket, a typed sister language of the Lisp dialect Racket, is able to statically type 
check most idiomatic Racket code.
These optional type systems have some important differences from standard static
type systems, for example they often include \emph{union} types.

Clojure is a recent and increasingly popular programming language that is designed to support 
functional programming and immutability by default. Implementations of Clojure are hosted 
on popular platforms and provide direct interoperability with the host platform.
There are several implementations of Clojure, each targeting different runtimes including
the Java Virtual Machine, the Common Language Runtime, and Javascript virtual machines.

This dissertation describes my work on designing an optional static type system for Clojure based on
the lessons learnt from several projects, primarily Typed Racket.
This work includes designing and building a type checker for Clojure running on the Java Virtual Machine.
Several experiments are conducted using this prototype, particularly involving
existing Clojure code that is sufficiently complicated that
type checking increases confidence that the code is correct.
For example, nearly all of \emph{algo.monads}, a Clojure Contrib library for monadic programming, 
is able to be type checked.
Most monad, monad transformer, and monadic function definitions can be type checked,
usually by adding type annotations in natural places like function definitions.

There is significant future work to fully type check all Clojure features and idioms.
For example, multimethod definitions and functions with particular constraints on the number of 
variable arguments they accept (particularly, functions taking only an even number of variable arguments) are troublesome. 
Also, there are desirable features from the Typed Racket project that are missing, such
as automatic runtime contract generation and a sophisticated blame system, 
both which are designed to improve error messages when mixing typed and untyped code in similar systems.
Overall, the work described in this dissertation leads to the conclusion that it appears to 
be both practical and useful to design and implement an optional static type system for the
Clojure programming language.

@@

Last November I attended my first conference, Clojure Conj 2011 (Raleigh, NC),
where I delivered a talk on logic programming with Clojure and met some of my programming heroes.
A week later my supervisor Rowan Davies invited me to visit Carnegie Mellon University
where he worked on his PhD project. Most memorably, I was invited to a meeting between
Rowan and his PhD supervisor Frank Pfenning.

@@

Many thanks go to my supervisor Rowan. His influence as a lecturer inspired my exploration of
functional programming languages. He has exceeded my expectations as an honours supervisor,
often dedicating time to assist me on public holidays and weekends.
Rowan's experience with refinement types brought a unique perspective to this project
which was very welcome.

I'd also like to thank: Sam Tobin-Hochstadt for his consistent assistance and patience;
David Nolen for mentoring this project as part of Google Summer of Code 2012;
and Chas Emerick, without whom I would have never considered attending Clojure Conj 2011
(and this work probably would not have happened).
Also, Daniel Spiewak, Dan Friedman and William Byrd, Jim Duey, Baishampayan Ghose, Sam Aaron, and
Steve Strickland, thank you for your help and inspiration.
Thank you Rich Hickey for creating Clojure, it is a thrill to build tools that make
this great language even better.

Fittingly, I have been invited to talk on Typed Clojure at Clojure Conj
2012 in November, which I am highly anticipating.

>>>>

\keywords{Programming languages, optional type systems, Clojure}
% TODO What are categories? cshonours requires them - but commented out now
%\categories{}

@@

%Traditionally there has been a clear distinction between 
%programming languages that perform type checking at compile-time, and those that do not.
%In particular, the early wide-spread programming languages ALGOL and Lisp
%could be divided this way: ALGOL performs compile-time (or static) type checking,
%and Lisp does not, and most recent languages can be similarly classified.
%Static type checking eliminates many common user errors that are otherwise unnoticed or are caught 
%when the program is run.
%On the other hand, a language without static type checking can often be designed 
%to be more flexible and forgiving, making it easy, for example, to have a \emph{map} function that can map functions
%with any number of arguments (as in most Lisp dialects).
%
%A recent trend is to aim to combine the advantages of both kinds of languages by adding optional static 
%type systems to languages without static type checking. 
%A major challenge in these combinations is continuing to support the idiomatic style of the original 
%language while performing accurate enough checking to catch similar errors to standard statically
%typed languages. Some recent attempts appear to have finally produced type systems that have overcome this obstacle.
%Notably Typed Racket, a typed sister language of the Lisp dialect Racket, is able to statically type 
%check most idiomatic Racket code.
%These optional type systems have some important differences from standard static
%type systems, for example they often include \emph{union} types.
%

Dynamic programming languages often abandon the advantages of static type checking 
in favour of their characteristic convenience and flexibility.
Static type checking eliminates many common user errors at compile-time that
are otherwise unnoticed, or are caught later in languages without static type checking.
A recent trend is to aim to combine the advantages of both kinds of languages by adding \emph{optional} static 
type systems to languages without static type checking, while preserving the languages idioms and style.

This dissertation describes my work on designing an optional static type system for the Clojure programming language,
a dynamically typed dialect of Lisp, based on
the lessons learnt from several projects, primarily Typed Racket.
This work includes designing and building a type checker for Clojure running on the Java Virtual Machine.
Several experiments are conducted using this prototype, particularly involving
existing Clojure code that is sufficiently complicated that
type checking increases confidence that the code is correct.
For example, nearly all of \emph{algo.monads}, a Clojure Contrib library for monadic programming, 
is able to be type checked.
Most monad, monad transformer, and monadic function definitions can be type checked,
usually by adding type annotations in natural places like function definitions.

There is significant future work to fully type check all Clojure features and idioms.
For example, multimethod definitions and functions with particular constraints on the number of 
variable arguments they accept (particularly functions taking only an even number of variable arguments) are troublesome. 
Also, there are desirable features from the Typed Racket project that are missing, such
as automatic runtime contract generation and a sophisticated blame system, 
both which are designed to improve error messages when mixing typed and untyped code in similar systems.

Overall, the work described in this dissertation leads to the conclusion that it appears to 
be both practical and useful to design and implement an optional static type system for the
Clojure programming language.

@@

Last November I attended my first programming conference, Clojure Conj 2011 (Raleigh, NC),
where I delivered a talk on logic programming with Clojure and met some of my programming heroes.
A week later my supervisor Rowan Davies invited me to visit Carnegie Mellon University
where he worked on his PhD project. Most memorably, I was invited to a meeting between
Rowan and his PhD supervisor Frank Pfenning.

@@

Many thanks go to my supervisor Rowan. His influence as a lecturer inspired my exploration of
functional programming languages. He has exceeded my expectations as an honours supervisor,
and we plan to continue to collaborate in this area.
Rowan's experience with refinement types brought a unique perspective to this project
which was very welcome.

I am grateful to Sam Tobin-Hochstadt for his pioneering work in this area of research.
He has always been happy to offer assistance and advise throughout this project, which always proved invaluable.

I would also like to thank David Nolen for mentoring this project as part of Google Summer of Code 2012,
and Chas Emerick, without whom I would have never even considered attending Clojure Conj 2011
(and this work probably would not have happened).
Also, Daniel Spiewak, Dan Friedman and William Byrd, Jim Duey, Baishampayan Ghose, Sam Aaron, and
Steve Strickland, thank you for your help and inspiration.
Thank you Rich Hickey for creating Clojure, it is a thrill to build tools that make
this great language even better.

Fittingly, I gave a talk on Typed Clojure at Clojure Conj 2012, which was extremely satisfying.
